Push-Pull Signal-Function Functional Reactive Programming

### 2012-11-29 Docs/IFL2012/abstract.tex
edits to ifl abstract and intro

\begin{abstract}
Functional Reactive Programming is a promising class of systems for writing
interactive and time-dependent programs. Signal-function FRP is a subclass of
these systems which provides advantages of modularity and correctness, but
has proven difficult to efficiently implement.

The abstraction of signal vectors provides the necessary type apparatus to
distinguish components of the input and output of signal functions which benefit
from a push-based implementation from those which benefit from a pull-based
implementation, and to combine both implementation strategies in a single system.

We describe a signal-function FRP system which provides push-based evaluation
for events, pull-based evaluation for signals, and a simple monadic evaluation
interface which permits the system to be easily integrated with one or more
IO systems.
\end{abstract}

>>>>

\begin{abstract}
Functional Reactive Programming is a promising model for writing interactive and
time-dependent programs. Signal-function FRP is a variant of FRP which provides
advantages in modularity and correctness, but has proven difficult to implement
efficiently.

One important implementation trade-off is whether evaluation of an FRP program
is push-based or pull-based. Signal vectors are an abstraction which addresses
this by providing a type-level description of the inputs and outputs of signal
functions, distinguishing components that benefit from a push-based
implementation from those that benefit from a pull-based implementation. This
makes it possible to combine both implementation strategies in a single system.

We describe a signal-function FRP system which provides push-based evaluation
for events, pull-based evaluation for signals, and a simple monadic evaluation
interface that permits the system to easily integrate with one or more IO
systems.
\end{abstract}

### 2012-11-29 Docs/IFL2012/introduction.tex
edits to 2.0
edits to ifl abstract and intro

Functional Reactive Programming (FRP) is a class of systems for describing
reactive programs. Reactive programs are programs which, rather than 
\new{batch-processing inputs to produce an output,}
% taking a single input and producing a single output, 
\new{instead map time-varying inputs onto outputs.}
%% must accept multiple inputs and 
%% alter temporal behavior, including the production of multiple outputs, based
%% on these inputs.

An FRP system provides a means of manipulating {\em behaviors} and
{\em events}. 
\rn{Switched to present tense.  Don't need to add additional future or past tense...}
Behaviors are often referred to as {\em signals} in FRP literature,
but the definition is the same.  Semantically, a behavior or signal is a
function from time to a value.  
\new{Whereas an event is a ordered sequence of discrete occurrences,
  carrying both values and timestamps.}
%% An event is a discrete, possibly infinite, and 
%% time-ordered sequence of occurrences, which are times paired with values.

\rn{Looks like you should standardize early on {\bf signal} and leave
  extra occurrences of {\bf behavior} out of this intro.}

FRP systems can generally be categorized as ``classic FRP,'' 
% which corresponds to the originally described FRP system 
in which behaviors
and events are manipulated directly as first-class values,
% in the FRP system,
or {\em signal-function FRP}. 
% in which behaviors  (generally termed signals in this approach) 
\new{In signal-function FRP, the programmer cannot directly manipulate
  signals, rather, functions from input-signal to output-signal are
  the central abstraction.  Thus}
%% and events are not first-class values, but signal
%% functions are first class values. 
signal functions are time-dependent,
reactive transformers of signals, events, or combinations of signals and events.
%
FRP {\em combines} behaviors and events through the use of {\em switching}, in which
a behavior (in classic FRP) or a signal function (in signal-function FRP) is
replaced by a new behavior or signal function carried by an event occurrence.

@@

Classic FRP was first described as a system for interactive animations~\cite{Elliott1997}.
Recent work on classic FRP has focused on efficient implementation \rn{CITATION?}.  One approach to
efficiency is to separate the evaluation of behaviors and events, since suitable 
strategies give best performance in each case. Push-based evaluation evaluates a
system only when input is available, and is thus suitable for discrete inputs
such as events. Pull-based evaluation evaluates the system as quickly as
possible, polling for input, and is preferable for behaviors and signals.
The initial implementations of FRP made use of pull-based evaluation for both
behaviors and events. ``Reactive''~\cite{Elliott2009}, as well as more recent systems such as
``reactive-banana''~\cite{Apfelmus}, make use of push-based evaluation for
events and pull-based evaluation for behaviors. This is known as ``push-pull''
evaluation.

All implementations of signal-function FRP to date~\cite{Courtney2001-1,Nilsson2002,Nilsson2005,Sculthorpe2011}
have used pull-based evaluation for both signals and events. This is due to
the ease of implementation of pull-based evaluation, and because the types used for
signal functions do not permit distinguishing signals and events, or
constructing only part of the input (for instance, one event occurrence).

\rn{What does it mean to construct one event occurrence?  As opposed to?}

@@

We present TimeFlies,\footnote{The sentence ``Time flies like an arrow.'' is a 
favorite quotation of one of the author's philosophy instructors, used to
demonstrate the ambiguity of language. The origin of the quotation is unknown.}
a push-pull signal-function FRP system. We hope to demonstrate the feasibility
of such an approach to FRP, and provide a basis for further research into
efficient implementation of signal-function FRP. We also describe a powerful
evaluation interface for TimeFlies, which permits us to use TimeFlies to
describe applications which make use of multiple and differing IO libraries.

\rn{Need a concrete contributions.   ``The first system to apply
  push-pull implementation strategies to signal-function FRP'', etc.
  It doesn't hurt for it to be bulletted.}

>>>>

Functional Reactive Programming (FRP) is a class of systems for describing
interactive programs. These are programs which, rather than batch-processing
inputs to produce an output, instead map time-varying inputs onto outputs.

An FRP system provides a means of manipulating {\em behaviors} and
{\em events}. Behaviors are often referred to as {\em signals} in FRP literature,
but the definition is the same.  Semantically, a behavior or signal is a
function from time to a value. An event is a ordered sequence of discrete
occurrences, carrying both values and timestamps.

FRP systems can generally be categorized as ``classic FRP,'' in which behaviors
and events are manipulated directly as first-class values,
or {\em signal-function FRP}. In signal-function FRP, the programmer cannot
directly manipulate signals, rather, the programmer manipulates time-dependent,
reactive transformers of signals, events, or combinations of signals and events.

FRP {\em combines} behaviors and events through the use of {\em switching}, in which
a behavior (in classic FRP) or a signal function (in signal-function FRP) is
replaced by a new behavior or signal function carried by an event occurrence.

@@

Classic FRP was first described as a system for interactive animations~\cite{Elliott1997}.
Recent work on classic FRP has focused on efficient implementation. One approach
to efficiency is to separate the evaluation of behaviors and events, since
suitable strategies give best performance in each case. Push-based evaluation
evaluates a system only when input is available, and is thus suitable for
discrete inputs such as events. Pull-based evaluation evaluates the system as
quickly as possible, polling for input, and is preferable for behaviors and
signals. The initial implementations of FRP made use of pull-based evaluation
for both behaviors and events. ``Reactive''~\cite{Elliott2009}, as well as more
recent systems such as ``reactive-banana''~\cite{Apfelmus}, make use of
push-based evaluation for events and pull-based evaluation for behaviors. This
is known as ``push-pull'' evaluation.

Almost all implementations of signal-function FRP to date~\cite{Courtney2001-1,Nilsson2002,Nilsson2005,Sculthorpe2011}
have used pull-based evaluation for both signals and events. This is due to the
ease of implementation of pull-based evaluation, and because the types used for
signal functions do not permit distinguishing signals and events, or
constructing only part of the input (for instance, an occurrence on one of the
event inputs). One signal function system~\cite{Jeffery2012} does permit the
user to push {\em segments} of signals, but, like the others, does not
distinguish events in a way that permits them to be separately pushed.

@@

TimeFlies,\footnote{The sentence ``Time flies like an arrow.'' is a 
favorite quotation of one of the author's philosophy instructors, used to
demonstrate the ambiguity of language. The origin of the quotation is unknown.}
is a push-pull signal-function FRP system, implemented in the Haskell
programming language. We show that typing signal-function FRP with signal
vectors allows push-pull evaluation of signal functions. We demonstrate an
evaluation interface that allows the user of TimeFlies to take advantage of this
behavior in interfacing TimeFlies to a wide range of IO systems. Finally, we
give benchmark results which demonstrate that the push-pull evaluation strategy
makes TimeFlies more efficient than the current pull-based implementation of
signal-function FRP.
